---
title: "Gov51 Project"
author: "Angela Calderon Pio"
format: pdf
---

# Preliminary RQ:

How is state-level per-pupil education spending in 1989-90 associated with intergenerational income mobility across U.S. commuting zones?

This research question has been refined from the initial proposal to better reflect the structure of the data. While the original question focused on state-level comparisons, the analysis uses commuting zones as the primary unit of observation, with state-level spending assigned to each zone.

Additionally, this analysis combines data from different sources that are measured at different points in time. Intergenerational mobility estimates are based on cohorts of children born in the early 1980s, while per-pupil education spending is drawn from the 1989-90 school year (which I changed from the time of my proposal). This year was selected because it aligns closely with the period during which these cohorts were in school, making it a more appropriate measure of the educational environment they experienced.

# Set-Up

```{r}

library(ggplot2)
library(readxl)
library(dplyr)
library(knitr)
library(readxl)
library(tidyr)


df1 <- readxl::read_excel(
  "/cloud/project/data/raw/online_data_tables-4.xls",
  sheet = 5,
  skip = 8
)

# Renaming column names because they didn't read in cleanly
df1 <- df1 |>
  rename(
    msa_id = `-1`,
    msa_name = `-2`,
    state = `-3`,
    n_children = `(4)`,
    rank_rank_slope = `(5)`,
    upward_mobility = `(6)`
  )

df1_state <- df1 |>
  group_by(state) |>
  summarize(mobility = mean(rank_rank_slope, na.rm = TRUE)) |>
  filter(!grepl("-", state))


df2 <- readxl::read_excel(
  "/cloud/project/data/raw/tabn194.xls",
  skip = 4,
  col_names = FALSE
)

df2 <- df2 |>
  select(state = 1, spending = 5) |>
  filter(!is.na(state), state != "United States") |>
  mutate(
    state = trimws(gsub("\\.+", "", state)),
    state = ifelse(state == "District of Columbia", "DC", state),
    state = ifelse(state == "DC", "DC", state.abb[match(state, state.name)]),
    spending = as.numeric(spending) 
  ) |>
  filter(!is.na(state))

df_merged <- df1 |>
  left_join(df2, by = "state") |>
  select(state, rank_rank_slope, spending)

all_states <- state.abb


```

# Summary Statistics

```{r}

df_merged |>
  summarize(
mean_spending = mean(spending, na.rm = TRUE),
    sd_spending = sd(spending, na.rm = TRUE),
    min_spending = min(spending, na.rm = TRUE),
    max_spending = max(spending, na.rm = TRUE),
    
    mean_mobility = mean(rank_rank_slope, na.rm = TRUE),
    sd_mobility = sd(rank_rank_slope, na.rm = TRUE),
    min_mobility = min(rank_rank_slope, na.rm = TRUE),
    max_mobility = max(rank_rank_slope, na.rm = TRUE),
    
    n = n()
  ) |>
  kable(
    digits = 2,
    col.names = c(
      "Mean Spending", "SD Spending", "Min Spending", "Max Spending",
      "Mean Mobility", "SD Mobility", "Min Mobility", "Max Mobility",
      "Observations"
    ),
    caption = "Summary Statistics (Table 1)"
  )

```

Table 1 presents summary statistics for per-pupil spending and intergenerational mobility across commuting zones. On average, per-pupil spending is approximately \$2,170.93, with variation (SD is about \$478.59), ranging from \$1,574.44 to \$4,727.59.

The mean rank-rank slope is 0.33, suggesting a moderate degree of persistence in income across generations, Mobility also varies across commuting zones, with values ranging from 0.18 to 0.43. Overall, these summary statistics indicate modest variation in both education spending and mobility.

# Figure 1

```{r}

northeast <- c("CT","ME","MA","NH","RI","VT","NJ","NY","PA")
midwest   <- c("IL","IN","MI","OH","WI","IA","KS","MN","MO","NE","ND","SD")
south     <- c("DE","FL","GA","MD","NC","SC","VA","DC","WV","AL","KY","MS","TN","AR","LA","OK","TX")
west      <- c("AK","AZ","CA","CO","HI","ID","MT","NV","NM","OR","UT","WA","WY")

df_merged <- df_merged |>
  mutate(
    region = case_when(
      state %in% northeast ~ "Northeast",
      state %in% midwest ~ "Midwest",
      state %in% south ~ "South",
      state %in% west ~ "West",
      TRUE ~ NA_character_
    )
  )

df_merged <- df_merged |>
  filter(!is.na(region))

ggplot(df_merged, aes(x = spending, y = rank_rank_slope, color = region)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Per-Pupil Spending",
    y = "Rank-Rank Slope (Higher = Less Mobility)",
    title = "School Spending and Intergenerational Mobility by Region"
  )
  
  model <- lm(rank_rank_slope ~ spending, data = df_merged)
summary(model)

```

> The scatterplot shows a weak overall relationship between per-pupil education spending and intergenerational mobility, with points widely dispersed across the range of spending levels. While some regional variation is visible, the trends are inconsistent: the South shows a slight negative relationship (higher spending associated with more mobility), while the Northeast and West exhibit slight positive trends (higher spending associated with less mobility). The Midwest is basically flat, indicating no "stickyness" between generations. These mixed patterns suggest that the relationship between spending and mobility is not consistent across regions and is generally weak.

\newpage

# Figure 2

```{r}

df_merged |>
  filter(!is.na(spending), !is.na(rank_rank_slope)) |>
  mutate(
    spending_group = ntile(spending, 4),
    spending_group = factor(
      spending_group,
      levels = c(1, 2, 3, 4),
      labels = c("Lowest spending", "Low-middle", "High-middle", "Highest spending")
    )
  ) |>
  group_by(spending_group) |>
  summarize(
    Mean = mean(rank_rank_slope, na.rm = TRUE),
    Median = median(rank_rank_slope, na.rm = TRUE),
    .groups = "drop"
  ) |>
  pivot_longer(
    cols = c(Mean, Median),
    names_to = "Statistic",
    values_to = "mobility"
  ) |>
  ggplot(aes(x = spending_group, y = mobility, fill = Statistic)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_fill_manual(values = c(
    "Mean" = "#b6a0e3",
    "Median" = "#4b3f72"
  )) +
  scale_x_discrete(labels = c(
    "Lowest\nspending",
    "Low-middle",
    "High-middle",
    "Highest\nspending"
  )) +
  labs(
    x = "Per-Pupil Spending Group",
    y = "Rank-Rank Slope\n(Higher = Less Mobility)",
    fill = "Statistic",
    title = "Mobility Patterns Are Uneven Across Spending Groups"
  )


```

Figure 2 groups commuting zones into four spending categories: lowest, low-middle, high-middle, and highest spending. Then, I compare their average rank-rank slopes. If higher spending were strongly associated with greater mobility, the bars would fall steadily as spending increases, since lower rank-rank slopes mean more mobility. Instead, the pattern is uneven. The high-middle spending group has the lowest average rank-rank slopes, but the highest spending group **does not** continue that decline. This suggests that spending is related to mobility only weakly and that higher spending alone is not associated with better mobility outcomes. I include the median because averages can be affected by skew of extreme values, so the median gives a better sense.

# Correlation

```{r}


cor(df_merged$spending, df_merged$rank_rank_slope, use = "complete.obs")

```

The correlation between per-pupil spending and the rank-rank slope is -0.135, indicating a very weak negative relationship. This suggests that higher spending is slightly associated with greater mobility (lower rank-rank slopes). However, the relationship remains small in magnitude, indicating that spending alone is not a strong predictor of mobility.

# Linear Regression

```{r}

model <- lm(rank_rank_slope ~ spending_1000, data = df_merged)

kable(
  summary(model)$coefficients,
  digits = 4,
  caption = "Linear Regression of Intergenerational Mobility on Per-Pupil Spending"
)
      
```

A simple linear regression shows a negative but small relationship between per-pupil spending and the rank-rank slope. Because spending is measured in thousands of dollars here, the coefficient means that each additional \$1,000 in per-pupil spending is associated with a 0.0158 decrease in the rank-rank slope. Since higher rank-rank slopes mean less mobility. This points in the expect direction: higher spending is associated with slightly greater mobility. However, the size of the coefficient is still modest, so the result should be interpreted as a weak association, not as strong evidence that spending itself explains mobility differences across commuting zones.
